\vfill World of Warcraft's New Way to Get You Hooked

### 2013-03-02 paper.tex
1.02 Analysis progress. Cut down paper to 2 SE Code sections.

% Two Column Format
\documentclass[11pt]{article}
%this allows us to specify sections to be single or multi column so that things like title page and table of contents are single column
\usepackage{multicol}

@@

% [5%] Abstract (75 Words Max)
% __ One or two sentences of context about your issue.
% __ Provide your focus question.
% __ Mention one or two relevant arguments.
% __ Briefly explain your conclusion and why it is proper ethically. (Optimal total is 7 sentences or so….)

\vfill  %in combination with \newpage this forces the abstract to the bottom of the page
\begin{abstract}
World of Warcraft uses a ``Starter Edition'' which allows players access for free indefinitely with restrictions. \cite{starterEditionNewsRelease} Studies show a correlation between time playing and addiction to Massively Multiplayer Online Role-Playing Games (MMORPGs) like World of Warcraft. \cite{ProblemUsageAmongHighlyEngagedPlayers}\cite{JustAComputerGameRight} Is it ethical for World of Warcraft to use this model?

Based on the principles of Utilitarianism \cite{Utilitarianism}, restricting a free demo on the basis of a few cases of mental instability is unethical, therefore World of Warcraft's ``Starter Edition'' \emph{is} ethical.
\end{abstract}

@@

World of Warcraft is an incredibly popular game with over 10 million subscribers. \cite{WoWPlayerCount} Players of World of Warcraft spend an average of 22 hours per week playing. \cite{PlayerMotivations} World of Warcraft is free to play provided the player is willing to accept some restrictions. None of those restrictions hinder the number of hours one can play. \cite{StarterEditionFAQ} Multiple studies have shown that time spent playing a MMORPG, like World of Warcraft, correlates to addiction level in some individuals. \cite{ProblemUsageAmongHighlyEngagedPlayers}\cite{JustAComputerGameRight} The intensity with which many players play World of Warcraft or other MMORPGs has led to multiple deaths. \cite{SnowlyDeath}\cite{ShawnDeath}

@@

\section{Research Question}
Is Blizzard's use of World of Warcraft: Starter Edition ethical? \\
World of Warcraft: Starter Edition could precipitate an ``outbreak'' of videogame addiction leading to terrible incidents. \cite{SnowlyDeath} Deaths and addiction linked to World of Warcraft would cause a sharp decline in the sale of games. Blizzard would suffer and be forced to lay off software engineers working on its products.

@@

\section{Arguments}
\subsection{Arguments Affirmative}
\subsubsection{Differing Player Archetypes}
Many studies and surveys disregard the fact that different types of people play MMORPGs for different reasons and these varying archetypes lead to a majority of players who are not in danger of addiction. By not making mention of this fact, sweeping generalizations are made without basis. \cite{PlayerMotivations}
\subsubsection{More Time Online Does Not Mean Addiction}
Many people who are characterized as spending ``excessive'' time online do not show symptoms of addiction, rather they show signs of psychological fulfillment that would otherwise elude them due to disabilities, circumstances, or social discomfort. Some people who are isolated or autistic find comfort in the use of the internet and games to create social relationships they would not have cultivated otherwise. \cite{IsThereEvidenceOfInternetAddiction}
\subsubsection{Excessive Users Do Not Exhibit Addiction Symptoms}
Users that spend more time playing the game than average exhibit characteristics that show using MMORPGs for social aspects of the in-game world as opposed to symptoms of addiction. These users would just as likely find fun elsewhere. \cite{ExcessiveUseForSocialAspects}
\subsection{Arguments Negative}
\subsubsection{More Intense Play By A Younger Audience Leads to Addiction}
``MMORPGs are played more intensely than video games,'' and are played primarily by a younger audience. Persons under the age of 27 (i.e. the majority of players of MMORPGs) are far more susceptible to addiction due to a lack of self identity and self confidence. \cite{YoungerAudiencesGetAddicted}
\subsubsection{Increased Time Playing Leads to Life Problems}
The longer a person plays World of Warcraft, the likelier they are to experience issues in their lives. \cite{ProblemUsageAmongHighlyEngagedPlayers}
\subsubsection{Addictive Games Have a Social Responsibility}
Current efforts to warn players of online games' addictiveness are too minimal and often non-existent. Warning labels should be printed on games and companies should contact users who play in excess of a set threshold. \cite{MMORPGSocialResponsibility}\\

Arguments in favor of my conclusion that World of Warcraft \emph{is} ethical boil down to the fact that people are complex and a simple correlation of time spent playing to a person being psychologically addicted is too far of a stretch. People play for a variety of reasons, many of which are psychologically fulfilling. Arguments against my conclusion bring up that often these players have immature brains and a lack of social identity due to age, leading to increased risk of addiction. They also cite studies that show increased play times leading to life issues and that the companies has a social responsibility to at least warn consumers of possible psychological consequences.

@@

Blizzard's World of Warcraft is a software system that runs on computers. \cite{WoWSystemReqs} Software systems are written by Software Engineers. \cite{ACMApprovesSECode} According to the Computer Society and the ACM, all Software Engineers should abide by the rules outlined in the the SE Code of Ethics. \cite{ACMApprovesSECode}\cite{SECode} Therefore, Blizzard's World of Warcraft Software Engineers are subject to "the Code." \cite{SECode}

\subsection{Definition of ``addiction''}
Because this paper will often reference psychological addiction, it is important to make sure this notion is concrete before continuing.
A person who is addicted will pathologically pursue rewards and/or relief through substance abuse or other behaviours, in this case playing World of Warcraft. \cite{DefinitionOfAddiction}

``Addiction is characterized by inability to consistently abstain, impairment in behavioral control, craving, diminished recognition of significant problems with one’s behaviors and interpersonal relationships, and a dysfunctional emotional response.'' \cite{DefinitionOfAddiction}

\subsection{SE Code Section 1.02\\Moderate the interests of the software engineer, the employer, the client and the users with the public good.}
Potential to argue that opponents of my position would say that the game leads to addiction and more time spent playing it means higher chances of addiction. Addiction being bad for the ``public good'' and \emph{not} in the ``interests of the client.'' I will respond with sources saying that time does not correlate to addiction and that high levels of play time do not mean addiction.
\subsection{SE Code Section 2.07\\Identify, document, and report significant issues of social concern, of which they are aware, in software or related documents, to the employer or the client.}
Opponents have suggested warning labels and/or contacting players who exhibit signs of addiction. I may have to concede this point, I see nothing wrong with a few added measures.
``Tip: Remember to take all things in moderation (even World of Warcraft!)'' - In-Game Tip! WoW is at least putting some effort forth.
\subsection{SE Code Section 2.09\\Promote no interest adverse to their employer or client, unless a higher ethical concern is being compromised; in that case, inform the employer or another appropriate authority of the ethical concern.}
The opposition will say that promoting more sales while increasing chances for game addiction is adverse to the client. Again I will argue that just because players are hooked and spending money, does not mean they are addicted and their lives are suffering.
\subsection{SE Code Section 4.04\\Not engage in deceptive financial practices such as bribery, double billing, or other improper financial practices.}
Getting young people hooked on games could be seen as deceptive financial practices in that young people lack impulse control and will be more willing to spend their money frivolously. I will argue that a person's money is their own business and it is not Blizzard's responsibility to ensure a consumer is spending their money wisely.

>>>>

% Two Column Format
\documentclass[11pt]{article}
%this allows us to specify sections to be single or multi column so that things like title page and table of contents are single column
\usepackage{multicol}
\usepackage{verbatim}

@@

% [5%] Abstract (75-100 Words Max)
% __ One or two sentences of context about your issue.
% __ Provide your focus question.
% __ Mention one or two relevant arguments.
% __ Briefly explain your conclusion and why it is proper ethically. (Optimal total is 7 sentences or so...)

\vfill  %in combination with \newpage this forces the abstract to the bottom of the page
\begin{abstract}
Blizzard Entertainment's World of Warcraft, an online videogame, has a free ``Starter Edition'' which allows players to access the game indefinitely with restrictions. \cite{starterEditionNewsRelease} Studies show a correlation between time spent playing and addiction to Massively Multiplayer Online Role-Playing Games (MMORPGs) like World of Warcraft. \cite{ProblemUsageAmongHighlyEngagedPlayers}\cite{JustAComputerGameRight} Is it ethical for World of Warcraft to use this model?

Based on the Software Engineering Code of Ethics \cite{SECode} Blizzard has acted ethically on all accounts and therefore World of Warcraft's ``Starter Edition'' \emph{is} ethical.
\end{abstract}

@@

Blizzard Entertainment's World of Warcraft is an incredibly popular game with over 10 million subscribers. \cite{WoWPlayerCount} Players of World of Warcraft spend an average of 22 hours per week playing. \cite{PlayerMotivations} World of Warcraft is free to play provided the player is willing to accept some restrictions. None of those restrictions hinder the number of hours one can play. \cite{StarterEditionFAQ} Multiple studies have shown that time spent playing a MMORPG, like World of Warcraft, correlates to addiction level in some individuals. \cite{ProblemUsageAmongHighlyEngagedPlayers}\cite{JustAComputerGameRight} The intensity with which many players play World of Warcraft or other MMORPGs has led to deaths. \cite{SnowlyDeath}\cite{ShawnDeath}

@@

\section{Research Question}
Is Blizzard's use of World of Warcraft: Starter Edition ethical? \\

World of Warcraft: Starter Edition could precipitate an ``outbreak'' of videogame addiction leading to terrible incidents. \cite{SnowlyDeath} Deaths and addiction linked to World of Warcraft would also cause a sharp decline in the sale of games. Blizzard made \$4.85 billion last year and has around 4700 employees. \cite{BlizzardFinances} \cite{BlizzardEmployeeCount} At its worst the company could fail and those 4700 employees would be without income and looking for work.

@@

\section{Arguments}
\subsection{Arguments Affirmative}
\subsubsection{Differing Player Archetypes}
Users play MMORPGs for a variety of reasons. Sweeping generalizations are often made without basis by grouping all players into a single type that is prone to addiction. \cite{PlayerMotivations}
\subsubsection{More Time Online Does Not Mean Addiction}
Many people who are characterized as spending ``excessive'' (not defined in the article) time online do not show symptoms of addiction. These individuals show signs of psychological fulfillment. Often these people would be unable to find such fulfillment elsewhere due to disabilities, circumstances, or social discomfort. Some people who are isolated or autistic find comfort in the use of the internet and games to create social relationships they would not have cultivated otherwise. \cite{IsThereEvidenceOfInternetAddiction}
\subsubsection{Excessive Users Do Not Exhibit Addiction Symptoms}
Users that spend more time than average playing the game exhibit social fulfillment. Many users who play more than average are playing MMORPGs for social aspects of the in-game world. These users would just as likely find fun elsewhere. \cite{ExcessiveUseForSocialAspects}
\subsection{Arguments Negative}
\subsubsection{More Intense Play By A Younger Audience Leads to Addiction}
``MMORPGs are played more intensely than video games,'' and are played primarily by a younger audience. Persons under the age of 27, the majority of players of MMORPGs, are far more susceptible to addiction due to a lack of self identity and self confidence. \cite{YoungerAudiencesGetAddicted}
\subsubsection{Increased Time Playing Leads to Life Problems}
The longer a person plays World of Warcraft, the more likely they are to experience issues in their lives. \cite{ProblemUsageAmongHighlyEngagedPlayers}
\subsubsection{Addictive Games Have a Social Responsibility}
Current efforts to warn players of online games' addictiveness are minimal and often non-existent. Warning labels outlining the possibility of addition should be printed on game boxes and companies should contact users who play in excess of a set threshold. \cite{MMORPGSocialResponsibility}\\

Arguments that World of Warcraft's model \emph{is} ethical boil down to the fact that people are complex and a simple correlation of time spent playing to a person being psychologically addicted is too far of a stretch. People play for a variety of reasons, many of which are psychologically fulfilling. Arguments that propose that World of Warcraft is causing addiction bring up that often players have immature brains and a lack of social identity due to age, leading to increased risk of addiction. They also cite studies that show increased play times leading to life issues and that the companies have social responsibility to at least warn consumers of possible psychological consequences.

@@

To aid in my analysis of whether or not Blizzard Entertainment's World of Warcraft trial model is ethical, the Software Engineering Code of Ethics (SE Code) will be used. The SE Code applies to ``professional software engineers, including practitioners, educators, managers, supervisors and policy makers, as well as trainees and students of the profession.'' \cite{SECode} Blizzard Entertainment's World of Warcraft is a software system that runs on computers. \cite{WoWSystemReqs} Software systems are written by software engineers. \cite{ACMApprovesSECode} According to the Computer Society and the Association for Computing Machinery (ACM), all professional software engineers must abide by the rules outlined in the the SE Code. \cite{ACMApprovesSECode}\cite{SECode} Blizzard Entertainment can be seen as a ``practitioner'' in that they are actively engaged in making software, in addition to some of its employees who are software engineers. Therefore, Blizzard Entertainment and Blizzard Entertainment's software engineers are subject to the SE Code. \cite{SECode} Being subject to the SE Code allows me to use it to conclude the ethics of World of Warcraft's trial model.

\subsection{Definition of ``addiction''}
Because this paper will often reference psychological addiction, it is important to make sure this notion is concrete before continuing.\\
A person who is addicted will pathologically pursue rewards and/or relief through substance abuse or other behaviors, in this case playing World of Warcraft. \cite{DefinitionOfAddiction}

Addiction can be seen as an \emph{inability to abstain} from something, \emph{impairment in behavior control} regarding this thing, and a \emph{craving} for this thing. There will be a diminished recognition of issues with one's behavior, suffering of interpersonal relationships, and a dysfunctional emotional response. \cite{DefinitionOfAddiction}

\subsection{SE Code Section 1.02\\Moderate the interests of the software engineer, the employer, the client and the users with the public good.}
In the next few sections this tenet is broken down into its parts. I will explore the ``interests of the software engineer,'' in this case World of Warcraft's software engineers, the ``interests of ... the employer,'' in this case Blizzard Entertainment, and lastly the ``interests of ... the client and the users,'' the player. Following that will be a discussion of the ``public good'' using Utilitarianism to guide my definition.\\
``World of Warcraft's Engineers shall moderate the interests of themselves, Blizzard Entertainment, and the players of World of Warcraft with the public good.''
\footnote{Ryan did a great job with his comments on the actual document.}
\subsubsection{Interests of the Software Engineer}
It is common sense that a software engineer strives to make working software. For World of Warcraft's software engineers, this success is measured by how many players enjoy new content and continue to play the game. Considerations are made for balance and gameplay as well as story. \cite{DevWatercooler} These engineers are also interested in keeping their jobs and continuing to make money. In order to keep their jobs, their interests must lie in getting players to come back to the game world often and play the game.
\subsubsection{Interests of Blizzard Entertainment}
Blizzard Entertainment is a publicly traded company. \cite{BlizzStock} As such its main goal is to show its shareholders that it is profiting and increases the volume of business it handles. In no uncertain terms, Blizzard Entertainment wants as many people to buy and play their games, World of Warcraft included, as possible. Since World of Warcraft is subscription-based \cite{WoWSubscription}, Blizzard Entertainment earns money every single month from World of Warcraft players. None of the other games that Blizzard makes are subscription based. \cite{BlizzGameList} Blizzard Entertainment's interests are in getting more players to buy their games, particularly World of Warcraft which nets them money every month.
\subsubsection{Interests of the Player}
By purchasing a game like World of Warcraft, players are committing themselves to pay Blizzard Entertainment every month. Some players are reaping benefits that they would not be able to get otherwise, such as an increase in social interactions. \cite{IsThereEvidenceOfInternetAddiction} One particular study found that MMORPG users play for a mixture of ten reasons which break down into three categories: achievement, social, and immersion. \cite{PlayerMotivations} (All information following in this section is from the same source.) MMORPGs offer a sense of achievement through completing quests, challenging others, and optimizing character performance. MMORPGs offer the chance to socialize with people you've never met, join groups to accomplish tasks, and form relationships with people through guilds and group events. MMORPGs also offer a sense of immersion through their expansive worlds that can be explored, the opportunity to role-play with others, and the ability to customize your characters in a nearly infinite set of ways. The interests of the player boil down to achievements, social interaction, and immersion in a new world.
\subsubsection{The ``Public Good''}
I define the ``public good'' to be a population that is not addicted to World of Warcraft and therefore a population that exhibits the ability to abstain from World of Warcraft, have unimpaired behavior control, and not crave playing World of Warcraft.

In order to argue that the ``public good'' is being upheld I assert that Utilitarianism is the most effective ethical system to use because it judges actions based on the net happiness of everyone. Utilitarianism is ``[t]he creed which ... holds that actions are right in proportion as they tend to promote happiness, wrong as they tend to produce the reverse of happiness.'' \cite{Utilitarianism}

World of Warcraft has over 10 million subscribers \cite{WoWPlayerCount} who average 22 hours of play time per week. \cite{PlayerMotivations} For those users that spend above average time, they are most likely just enjoying the social aspects of the game. \cite{ExcessiveUseForSocialAspects} These people would have fun doing other activities instead and would not be bothered by not being able to play the game for a day or two. \cite{ExcessiveUseForSocialAspects} PSYCHOLOGICAL FULLFILLMENT, OTHER ARGUMENTS, READ YOUR DAMN PAPERS AND GET MORE FACTS

The most shocking statistic came from Dr. Maressa Orzack, founder and coordinator of Computer Addiction Services \cite{CompAddictionServices}, who cites that up to 40\% of World of Warcraft players could be addicted. \cite{FortyPercentAddicted} If everyone in the world played World of Warcraft, a total of over 7 billion people \cite{WorldPopulation}, and Dr. Ozack's prediction is correct, 2.8 billion people would be addicted. Based purely on the numbers that leaves 60\% of the world population who are not addicted.

Having a majority of players who are \emph{not} addicted, and thus playing for pleasure or other purposes, is in line with the Utilitarian view of ethical and therefore in line with the ``public good.''

\subsubsection{Ethical Conclusion}
In order for Blizzard's actions to be ethical, its engineers must ``moderate the interests of themselves, Blizzard Entertainment, and the players of World of Warcraft with the public good.''
The public good is being upheld, Blizzard Entertainment's actions are ethical in accordance with SE Code section 1.02.

\begin{comment}
\subsection{SE Code Section 2.07\\Identify, document, and report significant issues of social concern, of which they are aware, in software or related documents, to the employer or the client.}
DICE Interview? Gamastura, sounded concerned.

Opponents have suggested warning labels and/or contacting players who exhibit signs of addiction. I may have to concede this point, I see nothing wrong with a few added measures.
``Tip: Remember to take all things in moderation (even World of Warcraft!)'' - In-Game Tip! WoW is at least putting some effort forth.

\end{comment}
